The role of cognateness in native spoken word recognition

### 2025-01-13 manuscript/manuscript.tex
Address NSG comments

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Cognateness and recognition of foreign
words }
\title{The role of cognateness in native spoken word recognition}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Gonzalo
Garcia-Castro}~\orcidlink{0000-0002-8553-4209}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\href{mailto:gonzalo.garciadecastro@upf.edu}{gonzalo.garciadecastro@upf.edu}\asep\textbf{Serene
Siow}~\orcidlink{0000-0001-6482-2191}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\href{mailto:siow.serene@gmail.com}{siow.serene@gmail.com}\asep\textbf{Kim
Plunkett}~\orcidlink{0000-0003-0216-7480}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\href{mailto:kim.plunkett@psy.ox.ac.uk}{kim.plunkett@psy.ox.ac.uk}\asep\textbf{Nuria
Sebastian-Galles}~\orcidlink{0000-0001-6938-2498}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\href{mailto:nuria.sebastian@upf.edu}{nuria.sebastian@upf.edu}}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Insert abstract here.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
cognate \sep spoken word
recognition \sep phonology \sep bilingualism \sep non-native speech
processing \sep 
lexical access

@@

When some German speakers listened to the song \emph{The Power} by SNAP!
(\emph{Coyote Ugly}, 2000), many of them misheard the line ``I've got
the power'' as \emph{Agatha Bauer}. When Spanish speakers listen to the
line ``Circumvent your thick ego'' from the song \emph{Pictures} by
System of a Down (\emph{Steal this Album!}, 2002), they tend to hear
\emph{Sácame de aquí, King-Kong} {[}take me out of here, King-Kong{]}.
Outrageous as these examples might sound, the reader may know a few
cases in their own native language. This auditory illusion is common
across languages and cultures
(\citeproc{ref-dembeck2015oberflachenubersetzung}{Dembeck, 2015};
\citeproc{ref-efimova2018homophonic}{Efimova et al., 2018}) and can feel
quite real, and often inevitable. In Japanese, this phenomenon takes the
name \emph{Soramimi} (lit. ``empty ear''). Soramimi is a particular case
of homophonic translation in which words or phrases that appear in the
lyrics of a song in one language are translated into similar-sounding
words and phrases in a different language, without necessarily
preserving the meaning (\citeproc{ref-gasparov2006semen}{Gasparov,
2006}). Homophonic translation also occurs outside the context of music.
For instance, it is often intentionally employed by translators to
preserve the meter and ``sound structure'' (e.g., rhymes) of an original
text in the language of translation (e.g.,
\citeproc{ref-levick2016translating}{Levick, 2016};
\citeproc{ref-pilshchikov2016semiotics}{Pilshchikov, 2016})\footnote{The
  rationale behind this translation approach is to ``sacrifice{[}s{]}
  fidelity to the image rather than the melodiousness of verse''
  (\citeproc{ref-briusov1911ot}{Briusov, 1911}). For instance, Edith
  Grossman's translation of \emph{El Quijote}
  (\citeproc{ref-cervantes}{Cervantes \& Grossman, 2002}) present a
  similar case, in which formal aspects of the original text in Spanish
  are favoured to some degree over equivalence to English, even
  sometimes keeping Spanish words from the original text in the English
  translation (e.g., \emph{Señor}, \emph{ínsula}).}. Although homophonic
translation is mostly seen as a perceptual curiosity or as a literary
figure, little attention has been paid to homophonic translation as a
spontanous psycholinguistic event.

Otake (\citeproc{ref-otake2007interlingual}{2007}) analysed 194
instances of Soramimi, broadcasted between 1992 and 2007 by the TV show
\emph{Soramimi hour}, in which examples of Soramimi were presented the
the audience for comedy purposes. These instances consisted of
homophonic translations of English song lyrics to words and phrases in
Japanese. The vast majority of the reported instances were phrases
(96\%) and few were single words (4\%). There was relatively high
variability in the degree to which phonetic features of the presented
English lyrics were preserved, with some resulting Japanese words or
phrases sharing high overlap with the original English strings of
sounds, and some sharing very little overlap. An analysis of the few
instances of homophonic translations of single words revealed three main
phonological processes that could explain how Japanese listeners
reconstructed the English input to generate the Japanese words:
insertion (e.g., \emph{cry} /\textipa{kraI}/ to \emph{kurai},
/\textipa{kuRai}/ {[}dark{]}), deletion (e.g., \emph{go},
/\textipa{goU}/ to \emph{go}, /go/ {[}go{]}, and alternation (e.g.,
\emph{low}, /\textipa{loU}/ to \emph{rou}, /\textipa{Roo}/ {[}wax{]}).
This suggests that the reported homophonic translations can be
explained, to some extent, as the result of the Japanese listeners
accommodating the strings of English sounds to the Japanese phoneme
inventory and Japanese phonotactics
(\citeproc{ref-dupoux1999epenthetic}{Dupoux et al., 1999};
\citeproc{ref-peperkamp2008perceptual}{Peperkamp et al., 2008}).

In the present work, we investigated the interplay between phonological
similarity and phonological neighbourhood density to examine how
listeners may activate lexical representations in their native language
when listening to an unfamiliar language. Although homophonic
translations do not necessarily take place with the preservation of
meaning---which results in outrageous mistranslations like the
aforementioned ones---homophonic translations sometimes may lead to
correct translations. For example, imagine an English native listening
to Dutch for the first time. This listener may encounter the word
/\textipa{"pINgUIn}/ (\emph{pinguin}), which sounds similar to its
English translation /\textipa{"pENgwIn}/ (penguin). Given enough
phonological similarity between the non-native word and its translation
in the native language, a naïve listener may succeed at word
comprehension when listening to an unfamiliar language. Such
phonological similarity between translation equivalents---known as
\emph{cognateness}---is common across many languages, and is often due
to typological closeness and/or socio-historical events involving the
speakers of these languages (e.g., migration, social contact). For
example, Romance languages such as Spanish and Catalan share many
form-similar translation equivalents
(\citeproc{ref-schepens2012distributions}{Schepens et al., 2012}), as in
the case of \emph{puerta} and \emph{porta} (\emph{door} in Spanish and
Catalan, respectively). Given no prior knowledge of either language, a
Spanish native speaker is likely to be much more successful at correctly
translating Catalan \emph{porta} than English \emph{door}, due to the
phonological and orthographic similarity of the former to the equivalent
in their native language.

When comparing the phonemic transcription of the French to that of its
Spanish translation /\textipa{"pweR.ta}/ (\emph{puerta}), one may be led
to infer that these two words do not have much overlap. But one thing to
note is that the voiced uvular fricative consonant /\textipa{K}/ and the
open-mid back rounded vowel /\textipa{O}/, which do not exist in Spanish
as phonemic categories, can be perceived as allophonic variations the
native (Spanish) phonemes /\textipa{R}/ and /o/, as described by
assimilation models of phoneme perception
(\citeproc{ref-best1988examination}{Best et al., 1988},
\citeproc{ref-best2001discrimination}{2001}). Although this phonetic
mismatch between the non-native word and its translation can have a
noticeable toll on comprehension Cutler et al.
(\citeproc{ref-cutler2004patterns}{2004}), the Spanish listener may
still be able to activate the phonological representation of
/\textipa{"pweR.ta}/ (\emph{puerta}) from hearing the word-form
/\textipa{"pOKt}/ (\emph{porte}), ultimately allowing them to access the
correct semantic representation. Native prosody also constrains the
occurrence of homophonic translations. For instance, Kentner
(\citeproc{ref-kentner2015rhythmic}{2015}) presented in the auditory
modality French-native and German-native participants with English songs
with lyrics. Homophonic translations followed language-specific
segmentation strategies. Most homophonic translations reported by
German-native participants resulted from inserting phrase and word
boundaries before stressed syllables (following the more frequent
trochaic pattern in German). On the contrary, most homophonic
translations reported by French-native participants resulted from
inserting phrase and word boundaries after stressed syllables, following
the more frequent iambic pattern in French.

To explore the lexical mechanisms triggered by unfamiliar speech
listening, we used a translation elicitation task in which participants
listened to words from an unfamiliar language, and then tried to guess
the words translations in their native language. We will henceforth
refer to the auditory-presented words heard by participants on each
trial as \emph{presented words}, and the correct translation for the
presented words as \emph{target translations}. By manipulating the
amount of phonological similarity between the presented words and their
target translations, we explored listeners' reliance on phonological
similarity to succeed in the task. Since participants were unfamiliar
with the presented language, they could only use phonological similarity
between the presented language and their native language to successfully
translate the words. We therefore predicted that participants'
performance should increase when the translation pairs are
phonologically more similar. We also predicted that there is a minimum
threshold of phonological similarity to be sufficient for translation.

@@

For each word in the Catalan and Spanish lists pair, we defined three
predictors of interest: the lexical frequency of the correct English
translation (\emph{Frequency}), the phonological similarity between the
presented word and their correct English translation
(\emph{Similarity}), and the presented word's number of cross-language
phonological neighbours (\emph{CLPN}) in English. \emph{Frequency} was
included as a nuance predictor, under the hypothesis that---keeping
other predictors constant---participants would provide correct responses
more often when the correct translations are high frequency words in
their native language. Lexical frequencies of correct translations were
extracted from SUBTLEX-UK (\citeproc{ref-van2014subtlex}{Van Heuven et
al., 2014}), and transformed to Zipf scores. Translations without a
lexical frequency value were excluded from data analysis (2 in the
Catalan list, 5 in the Spanish list).

We calculated \emph{Similarity}, our main predictor of interest, by
computing the Levenshtein similarity between the X-SAMPA transcriptions
of each pair of translations using the \texttt{stringdist} R package
(\citeproc{ref-van2014stringdist}{van der Loo, 2014}). The Levenshtein
distance computes the edit distance between two character strings---in
this case, two phonological transcriptions---by counting the number of
additions, deletions, and substitutions necessary to make both strings
identical (\citeproc{ref-levenshtein1966binary}{Levenshtein, 1966}). We
divided this edit distance by the number of characters in the longest
X-SAMPA transcription of the translation pair. This transformation
accounts for the fact that Levenshtein distance tends to increase with
the length of the transcriptions. For interpretability, we subtracted
this proportion from one, so that values closer to one correspond to
higher similarity between phonological transcriptions, and values closer
to zero correspond to lower similarity between phonological
transcriptions. For example, the \emph{table}
(\texttt{teIb@l})-\emph{mesa} (\texttt{mesa}) translation pair had 17\%
similarity, while the \emph{train} (\texttt{trEIn})-\emph{tren}
(\texttt{tREn}) translation pair had 60\% similarity.

For each Catalan and Spanish word, we calculated the number of
\emph{CLPN} by counting the number of English words with same or higher
lexical frequency, and whose phonological transcription (in X-SAMPA
format) differed in up to one phoneme from that of the presented Catalan
or Spanish word. Lexical frequencies and phonological transcriptions
were extracted from the multilingual database CLEARPOND
(\citeproc{ref-marian2012clearpond}{Marian et al., 2012})\footnote{{[}Phonological
  transcriptions in CLEARPOND were generated from eSPEAK,
  \url{http://espeak.sourceforge.net/}{]}}. Table~\ref{tbl-stimuli}
summarises the lexical frequency, phonological neighbourhood density and
phonological overlap of the words included in the Catalan and the
Spanish lists.

@@

We modelled the probability of participants guessing the correct
translation of each presented word using a generalised multilevel
Bayesian regression model with a Bernoulli logit link distribution. We
included as fixed effects the intercept, the main effects of
\emph{Frequency}, \emph{Similarity}, and \emph{CLPN}, and the two-way
interaction between \emph{Similarity} and \emph{CLPN}. We also included
participant-level random intercepts and slopes for the main effects and
the interaction. Eq. 1 shows a formal description of the model.

\begin{equation}\phantomsection\label{eq-1}{
\begin{aligned}
&\textbf{Likelihood}  \\
y_{i} \sim & \text{Bernoulli}(p_{i}) \\ \\
&\textbf{Parameters}  \\
\text{Logit}(p_{i}) = &  \beta_{0[p,w]} + \beta_{1[p]} \text{Frequency}_{i} + \beta_{2[p]} \text{PTHN}_i + \beta_{3[p]} \text{Similarity}_i + \beta_{4[p]} (\text{PTHN}_i \times \text{Similarity}_i) \\
\beta_{0-6[p,w]} \sim & \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \text{ and  word } w \text{ in 1, ..., } W \\
\beta_{1-6[p]} \sim &  \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \\ \\
&\textbf{Prior}  \\
\mu_{\beta_{p,w}}  \sim &  \mathcal{N}(0, 0.1) \\
\sigma_{\beta_{p}},  \sigma_{\beta_{w}} \sim & \text{HalfCauchy}(0, 0.1) \\
\rho_{p}, \rho_{w} \sim & \text{LKJCorr}(8) \\
\end{aligned}
}\end{equation}

To test the practical relevance of each predictor we followed Kruschke
\& Liddell (\citeproc{ref-kruschke2018bayesian}{2018}). We first
specified a region of practical equivalence (ROPE) around zero ({[}-0.1,
+0.1{]}, in the logit scale). This area indicates the values of the
regression coefficients that we considered as equivalent to zero. We
then computed the 95\% posterior credible intervals (CrI) of the
regression coefficients of interest. Finally, we calculated the
proportion of the 95\% CrI that fell inside the ROPE. This proportion
indicates the probability that the true value of the coefficient is
equivalent to zero. All analyses were performed in R environment
(\citeproc{ref-rcoreteam2013language}{R Core Team, 2013}). We used the
tidyverse family of R packages
(\citeproc{ref-wickham2019welcome}{Wickham et al., 2019}) to process
data and to generate figures. We used the \texttt{brms} R package
(\citeproc{ref-burkner2017brms}{Bürkner, 2017}) using the
\texttt{cmdstanr} back-end to the Stan probabilistic language
(\citeproc{ref-carpenter2017stan}{Carpenter et al., 2017}) to estimate
and compare the models (see Appendix 1 for mode details on the models).

@@

\caption{\label{tbl-dataset}}

@@

Stimuli were the same list of Catalan stimuli as in Experiment 1.
Procedure and data analysis were identical as in Experiment 1.

@@

Experiment 2 was an extension of Experiment 1 to a population of
monolinguals whose native language is typologically close to the
presented language. We presented Catalan words to Spanish native adults
who were reportedly unfamiliar with Catalan. Our results indicate a
similar pattern of results as those in Experiment 1: participants were
able to provide correct translations of presented Catalan words,
provided that the Catalan words shared some degree of phonological
similarity with their Spanish translation, and that the number of
phonological neighbours with higher lexical frequency was reduced. In
contrast with the results in Experiment 1, the positive impact of
phonological similarity on participants' performance in Experiment 2 was
more resilient to the interference of phonological neighbourhood size.
English natives in Experiment 1 provided significantly less accurate
responses when more than four phonological neighbors were present (even
when translating high-similarity words), compared to when less than
three neighbours were present. Spanish participants in Experiment 2
benefited from phonological similarity even when eight neighbours were
present. Spanish participants' performance declined after eight
neighbours, and was most evident at 12 neighbours. Overall, this
suggests that participants in Experiment 2, who were natives of a
typologically similar language (Spanish) to the presented language
(Catalan), benefited more strongly from phonological similarity than
participants in Experiment 1, who were natives of typologically less
similar language (English) to the presented language (Catalan, Spanish).

Participants from both Experiment 1 and 2 benefited strongly from
phonological similarity to correctly translate words from a non-native,
reportedly unfamiliar language. This pattern of results holds for most
of the presented stimuli, but some low-similarity Catalan and Spanish
words were responded to surprisingly accurately by English listeners.
Given that participants were reportedly unfamiliar with both languages,
it was expected that participants would be very unlikely to provide
correct translations for words sharing little to no phonological
similarity to their correct translation. Table~\ref{tbl-surprises} shows
a list of Catalan and Spanish words to which participants provided
responses with \(\geq\) 10 average accuracy.

>>>>

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Cognateness and recognition of foreign
words }
\title{The role of cognateness in native spoken word
recognition\thanks{The authors declare no conflicts of interest with
regard to the funding source of this study. This study was supported by
the Spanish Ministry for Science and Innovation and State Research
Agency (Project PID2021- 123416NB-I00 financed by MCIN/ AEI/
10.13039/501100011033 / FEDER, UE) and the Economic and Social Research
Council (ESRC) (ES/S010947/1, UK). GGC was supported by a FPI research
contract (PRE2019-088165). We are grateful to Irina Lepadatu and Nicola
Gillen, and to the Speech Acquisition and Perception research group for
their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, Pamela
Miller, and Cristina Cuadrado for their technical support. We also thank
all families and infants who participated in the experiments. This study
was conducted according to guidelines laid down in the Declaration of
Helsinki, and was approved by the Drug Research Ethical Committee (CEIm)
of the IMIM Parc de Salut Mar, reference 2020/9080/I. The data and code
necessary to reproduce the analyses presented here are publicly
accessible, as are the materials necessary to attempt to replicate the
findings. The analyses presented here were not preregistered. Data,
code, and materials for this research are available at the following
URLs: https://osf.io/hy984/,
https://github.com/gongcastro/translation-elicitation.}}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Gonzalo
Garcia-Castro}~\orcidlink{0000-0002-8553-4209}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\href{mailto:gonzalo.garciadecastro@upf.edu}{gonzalo.garciadecastro@upf.edu}\asep\textbf{Serene
Siow}~\orcidlink{0000-0001-6482-2191}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\href{mailto:siow.serene@gmail.com}{siow.serene@gmail.com}\asep\textbf{Kim
Plunkett}~\orcidlink{0000-0003-0216-7480}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\href{mailto:kim.plunkett@psy.ox.ac.uk}{kim.plunkett@psy.ox.ac.uk}\asep\textbf{Nuria
Sebastian-Galles}~\orcidlink{0000-0001-6938-2498}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\href{mailto:nuria.sebastian@upf.edu}{nuria.sebastian@upf.edu}}
\date{}
\begin{document}
\maketitle
\begin{abstract}
While listening to speech in an unfamiliar language, words and phrases
in the native language are activated, leading to common auditory
illusions like Soramimi, a particular case of homophonic translation.
Homophonic translation has received little attention as a
psycholinguistic phenomenon. In this study, we explored the mechanisms
underlying lexical activation and selection in homoponic translation.
Across three studies, English and Spanish native adults completed a
translation elicitation task, in which they listened to a series of
Catalan or Spanish words (to which participants were unfamiliar), and
then had to type their best-guess translation in their native language.
Both English and Spanish natives were surprisingly good at translating
unfamiliar words, efficiently exploiting the phonological similarity
between the presented (unfamiliar) words and their correct translation.
When the correct translation belonged to high-density phonological
neighbours, participants' ability to benefit from phonological
similarity decreased. Finally, Spanish natives exploited phonological
similarity more efficiently: provided identical phonological similarity,
Spanish natives were more likely to guess the correct translation of the
unfamiliar word than English natives. Overall, we show that speech in an
unfamiliar language triggers equivalent dynamics of lexical selection
than native speech, providing a psycholinguistic account for homophonic
translation.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
cognate \sep spoken word
recognition \sep phonology \sep bilingualism \sep non-native speech
processing \sep 
lexical access

@@

When some German speakers listened to the song \emph{The Power} by SNAP!
(\emph{Coyote Ugly}, 2000), many of them misheard the line ``I've got
the power'' as \emph{Agatha Bauer}. When Spanish speakers listen to the
line ``Circumvent your thick ego'' from the song \emph{Pictures} by
System of a Down (\emph{Steal this Album!}, 2002), they tend to hear
\emph{Sácame de aquí, King-Kong} {[}take me out of here, King-Kong{]}.
Outrageous as these examples might sound, the reader may know a few
cases in their own native language. This auditory illusion is common
across languages and cultures and can feel quite real, often inevitable
(\citeproc{ref-dembeck2015oberflachenubersetzung}{Dembeck, 2015};
\citeproc{ref-efimova2018homophonic}{Efimova et al., 2018}). In
Japanese, this phenomenon takes the name \emph{Soramimi} (lit. ``empty
ear''). From a linguistic point of view, Soramimi is a particular case
of homophonic translation: words or phrases that appear in the speech
stream in one language are translated into similar-sounding words and
phrases in a different language, without necessarily preserving the
meaning (\citeproc{ref-gasparov2006semen}{Gasparov, 2006}). Homophonic
translation has mostly received attention as a perceptual curiosity, and
as a literary figure, intentionally employed by translators to preserve
the meter and ``sound structure'' (e.g., rhymes) of an original text in
the language of translation (e.g.,
\citeproc{ref-levick2016translating}{Levick, 2016};
\citeproc{ref-pilshchikov2016semiotics}{Pilshchikov, 2016})\footnote{The
  rationale behind this translation approach is to ``sacrifice{[}s{]}
  fidelity to the image rather than the melodiousness of verse''
  (\citeproc{ref-briusov1911ot}{Briusov, 1911}). For instance, Edith
  Grossman's translation of \emph{El Quijote}
  (\citeproc{ref-cervantes}{Cervantes \& Grossman, 2002}) present a
  similar case, in which formal aspects of the original text in Spanish
  are favoured to some degree over equivalence to English, even
  sometimes keeping Spanish words from the original text in the English
  translation (e.g., \emph{Señor}, \emph{ínsula}).}. Little attention
has been paid to homophonic translation as a psycholinguistic event.

In one of the few available studies concerning homophonic translation as
a psycholinguistic phenomenon, Otake
(\citeproc{ref-otake2007interlingual}{2007}) analysed 194 instances of
Soramimi, broadcasted between 1992 and 2007 by the Japanese TV show
\emph{Soramimi hour}, in which examples of Soramimi were presented to
the audience for comedy purposes. These instances consisted of
homophonic translations of English song lyrics to words and phrases in
Japanese. The vast majority of the reported instances were phrases
(96\%) and few were single words (4\%). There was relatively high
variability in the degree to which phonetic features of the presented
English lyrics were preserved, with some resulting Japanese words or
phrases sharing high overlap with the original English strings of
sounds, and some sharing very little overlap. An analysis of the few
instances of homophonic translations of single words revealed three main
phonological processes that could explain how Japanese listeners
reconstructed the English input to generate the Japanese words:
insertion (e.g., \emph{cry} /\textipa{kraI}/ to \emph{kurai},
/\textipa{kuRai}/ {[}dark{]}), deletion (e.g., \emph{go},
/\textipa{goU}/ to \emph{go}, /go/ {[}go{]}, and alternation (e.g.,
\emph{low}, /\textipa{loU}/ to \emph{rou}, /\textipa{Roo}/ {[}wax{]}).
This suggests that the reported homophonic translations can be
explained, to some extent, as the result of the Japanese listeners
accommodating the strings of English sounds to the Japanese phoneme
inventory and Japanese phonotactics
(\citeproc{ref-dupoux1999epenthetic}{Dupoux et al., 1999};
\citeproc{ref-peperkamp2008perceptual}{Peperkamp et al., 2008}). The
phonological processes described by Otake
(\citeproc{ref-otake2007interlingual}{2007}) do not take place in a
vaccum, but rather in the context of a lexicon. As words or phrases in
the native lexicon are activated by their acoustic similarity with the
non-native speech stream, they trigger a series of mechanisms that
result in the selection of one of the candidate word or phrase in the
native language. In the present work, we investigated the interplay
between phonological similarity and one of the main mechanisms involved
in the dynamics of lexical selection: phonological neighbourhood
density. The phonological neighbourhood density of a given word-form
refers to the number of words-forms that differ in a single phoneme from
the former. For instance, the English word \emph{cat} hbelongs to a
dense phonological neighbourhood, with 50 English phonological
neighbours (e.g., hat, bat, cap), while the word \emph{charger} belongs
to a sparse phonological neighbourhood, with six neighbours (e.g.,
charged,larger, charter, charmer). Phonological neighbourhood density
affects lexical processing. For instance, words from dense
neighbourhoods are recognized more slowly and less accuractely in the
auditory modality than words from sparse neighbourhoods
(\citeproc{ref-dufour2010phonological}{Dufour \& Frauenfelder, 2010};
e.g., \citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998};
\citeproc{ref-vitevitch2006clustering}{M. Vitevitch, 2006};
\citeproc{ref-vitevitch1998words}{Vitevitch \& Luce, 1998}). In this
study, we examined how listeners may activate lexical representations in
their native language when listening to an unfamiliar language.

Although homophonic translations do not necessarily take place with the
preservation of meaning---which results in outrageous mistranslations
like the aforementioned ones---homophonic translations sometimes may
lead to correct translations. For example, imagine an English native
listening to Dutch for the first time. This listener may encounter the
word /\textipa{"pINgUIn}/ (\emph{pinguin}), which sounds similar to its
English translation /\textipa{"pENgwIn}/ (penguin). Given enough
phonological similarity between the non-native word and its translation
in the native language, a naïve listener may succeed at word
comprehension when listening to an unfamiliar language. Such
phonological similarity between translation equivalents---known as
\emph{cognateness}---is common across many languages, and is often due
to typological closeness and/or socio-historical events involving the
speakers of these languages (e.g., migration, social contact). For
example, Romance languages such as Spanish and Catalan share many
form-similar translation equivalents
(\citeproc{ref-schepens2012distributions}{Schepens et al., 2012}), as in
the case of \emph{puerta} and \emph{porta} (\emph{door} in Spanish and
Catalan, respectively). Given no prior knowledge of either language, a
Spanish native speaker is likely to be much more successful at correctly
translating Catalan \emph{porta} than English \emph{door}, due to the
phonological and (orthographic similarity) of the former to the
equivalent in their native language.

To explore the lexical mechanisms triggered by unfamiliar speech
listening, we used a translation elicitation task in which participants
listened to words from an unfamiliar language, and then tried to guess
the words translations in their native language. We will henceforth
refer to the auditory-presented words heard by participants on each
trial as \emph{presented words}, and the correct translation for the
presented words as \emph{target translations}. By manipulating the
amount of phonological similarity between the presented words and their
target translations, we explored listeners' reliance on phonological
similarity to succeed in the task. Since participants were unfamiliar
with the presented language, they could only use phonological similarity
between the presented language and their native language to successfully
translate the words. We therefore predicted that participants'
performance should increase when the translation pairs are
phonologically more similar. We also predicted that there is a minimum
threshold of phonological similarity to be sufficient for correct
translation.

@@

For each word pair in the Catalan and Spanish lists, we defined three
predictors of interest: the lexical frequency of the correct English
translation (\emph{Frequency}), the phonological similarity between the
presented word and their correct English translation
(\emph{Similarity}), and the presented word's number of cross-language
phonological neighbours (\emph{CLPN}) in English. \emph{Frequency} was
included as a nuance predictor, under the hypothesis that---keeping
other predictors constant---participants would provide correct responses
more often when the correct translations are high frequency words in
their native language. Lexical frequencies of correct translations were
extracted from SUBTLEX-UK (\citeproc{ref-van2014subtlex}{Van Heuven et
al., 2014}), and transformed to Zipf scores. Translations without a
lexical frequency value were excluded from data analysis (2 in the
Catalan list, 5 in the Spanish list). \emph{Similarity}, our main
predictor of interest, was calculated as the Levenshtein similarity
between the X-SAMPA transcriptions of each pair of translations using
the \texttt{stringdist} R package (\citeproc{ref-van2014stringdist}{van
der Loo, 2014}). The Levenshtein distance computes the edit distance
between two character strings---in this case, two phonological
transcriptions---by counting the number of additions, deletions, and
substitutions necessary to make both strings identical
(\citeproc{ref-levenshtein1966binary}{Levenshtein, 1966}). We divided
this edit distance by the number of characters in the longest X-SAMPA
transcription of the translation pair. This transformation accounts for
the fact that Levenshtein distance tends to increase with the length of
the transcriptions. For interpretability, we subtracted this proportion
from one, so that values closer to one correspond to higher similarity
between phonological transcriptions, and values closer to zero
correspond to lower similarity between phonological transcriptions. For
example, the \emph{table} (\texttt{teIb@l})-\emph{mesa} (\texttt{mesa})
translation pair had 17\% similarity, while the \emph{train}
(\texttt{trEIn})-\emph{tren} (\texttt{tREn}) translation pair had 60\%
similarity. We calculated the number of \emph{CLPN} for each Catalan and
Spanish word by counting the number of English words with same or higher
lexical frequency, and whose phonological transcription (in X-SAMPA
format) differed in up to one phoneme from that of the presented Catalan
or Spanish word. Lexical frequencies and phonological transcriptions
were extracted from the multilingual database CLEARPOND
(\citeproc{ref-marian2012clearpond}{Marian et al., 2012})\footnote{{[}Phonological
  transcriptions in CLEARPOND were generated from eSPEAK,
  \url{http://espeak.sourceforge.net/}{]}}. Table~\ref{tbl-stimuli}
summarises the lexical frequency, phonological neighbourhood density and
phonological overlap of the words included in the Catalan and the
Spanish lists.

@@

We modelled the probability of participants guessing the correct
translation of each presented word using a generalised multilevel
Bayesian regression model with a Bernoulli logit link distribution. We
included as fixed effects the intercept, the main effects of
\emph{Frequency}, \emph{Similarity}, \emph{CLPN}, and \emph{Group}
(sum-coded as \texttt{cat-ENG\ =\ -0}, \texttt{spa-ENG\ =\ +0.5},
\citeproc{ref-schad2020capitalize}{Schad et al., 2020}) and the
three-way interaction between \emph{Similarity}, \emph{CLPN}, and
\emph{Group}. The predictor \emph{Group} was included to account for any
differences in task performance between English participants translating
Catalan and Spanish words. We also included participant-level random
intercepts and slopes for the main effects and the interaction. Eq. 1
shows a formal description of the model.

\begin{equation}\phantomsection\label{eq-1}{
\begin{aligned}
&\textbf{Likelihood}  \\
y_{i} \sim & \text{Bernoulli}(p_{i}) \\ \\
&\textbf{Parameters}  \\
\text{Logit}(p_{i}) = &  \beta_{0[p,w]} + \beta_{1[p]} \text{Frequency}_{i} + \beta_{2[p]} \text{Similarity}_i +  \beta_{3[p]} \text{CLPN}_i + \beta_{4[p]} \text{Group}_i + \\
&\beta_{5[p]} (\text{Similarity}_i \times \text{CLPN}_i) + \beta_{6[p]} (\text{Similarity}_i \times \text{Group}_i) + \\
& \beta_{7[p]} (\text{CLPN}_i \times \text{Group}_i) + \beta_{8[p]} (\text{Similarity}_i \times \text{CLPN}_i \times \text{Group}_i) \\ \\
\beta_{0-8[p,w]} \sim & \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \text{ and  word } w \text{ in 1, ..., } W \\
\beta_{1-8[p]} \sim &  \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \\ \\
&\textbf{Prior}  \\
\mu_{\beta_{p,w}}  \sim &  \mathcal{N}(0, 0.1) \\
\sigma_{\beta_{p}},  \sigma_{\beta_{w}} \sim & \text{HalfCauchy}(0, 0.1) \\
\rho_{p}, \rho_{w} \sim & \text{LKJCorr}(8) \\
\end{aligned}
}\end{equation}

To test the practical relevance of each predictor we followed Kruschke
\& Liddell (\citeproc{ref-kruschke2018bayesian}{2018}). We first
specified a region of practical equivalence (ROPE) around zero ({[}-0.1,
+0.1{]}, in the logit scale). This area indicates the values of the
regression coefficients that we considered as equivalent to zero. We
then computed the 95\% posterior credible intervals (CrI) of the
regression coefficients of interest. Finally, we calculated the
proportion of the 95\% CrI that fell inside the ROPE. This proportion
indicates the probability that the true value of the coefficient is
equivalent to zero. All analyses were performed in R environment
(\citeproc{ref-rcoreteam2013language}{R Core Team, 2013}). We used the
tidyverse family of R packages
(\citeproc{ref-wickham2019welcome}{Wickham et al., 2019}) to process
data and to generate figures. We used the \texttt{brms} R package
(\citeproc{ref-burkner2017brms}{Bürkner, 2017}) using the
\texttt{cmdstanr} back-end to the Stan probabilistic language
(\citeproc{ref-carpenter2017stan}{Carpenter et al., 2017}) to estimate
and compare the models (see Appendix 1 for mode details on the models).
Numeric predictors were standardised before entering the model by
subtracting the mean and dividing by the standard deviation. Categorical
predictors were sum-coded as \texttt{-0.5} and \texttt{+0.5}
(\citeproc{ref-schad2020capitalize}{Schad et al., 2020}).

@@

\caption{\label{tbl-dataset}Summary of participants' accuracy in the
translation elicitation task across Experiments 1, 2, and 3.}

@@

Stimuli were the same list of Catalan stimuli as in Experiment 1.
Procedure and data analysis were identical as in Experiment 1, with the
only exception that the statistical model did not include the
\emph{Group} predictor, given that only one group (\texttt{cat-SPA})
participated in this experiment.

@@

Experiment 2 was an extension of Experiment 1 to a population of
monolinguals whose native language is typologically close to the
presented language. We presented Catalan words to Spanish native adults
who were reportedly unfamiliar with Catalan. Our results indicate a
similar pattern of results as those in Experiment 1: participants were
able to provide correct translations of presented Catalan words,
provided that the Catalan words shared some degree of phonological
similarity with their Spanish translation, and that the number of
phonological neighbours with higher lexical frequency was reduced. In
contrast with the results in Experiment 1, the positive impact of
phonological similarity on participants' performance in Experiment 2
larger. Spanish natives in Experiment 2 provided exploited phonological
similarity more strongly than English natives in Experiment 1. Overall,
this suggests that participants in Experiment 2, who were natives of a
typologically similar language (Spanish) to the presented language
(Catalan), benefited more strongly from phonological similarity than
participants in Experiment 1, who were natives of typologically less
similar language (English) to the presented language (Catalan, Spanish).
Participants from both Experiment 1 and 2 benefited strongly from
phonological similarity to correctly translate words from a non-native,
reportedly unfamiliar language. This pattern of results holds for most
of the presented stimuli, but some low-similarity Catalan and Spanish
words were responded to surprisingly accurately by English listeners.
Given that participants were reportedly unfamiliar with both languages,
it was expected that participants would be very unlikely to provide
correct translations for words sharing little to no phonological
similarity to their correct translation. Table~\ref{tbl-surprises} shows
a list of Catalan and Spanish words to which participants provided
responses with \(\geq\) 10 average accuracy.
